---
title: "How Consequentialist Reasoning Shapes Bioethical Views"
author: "Piotr Bystranowski, Jon Rueda, & Ivar R. Hannikainen"
date: "8 02 2022"
output:
  github_document:
  #pdf_document:
  #  toc: yes
  html_document:
    code_folding: hide
    df_print: kable
    fig_height: 4
    highlight: monochrome
    theme: sandstone
    toc: yes
    toc_float: yes
---

```{r include = FALSE}
library(tidyverse)
library(flextable)

knitr::opts_chunk$set(echo=FALSE, warning = F, message = F)
```

```{r}
pretest <- read.csv("data/00_long_data.csv")
source('src/tables.R')
```

We initially drafted 14 normative and 16 factual statements regarding eleven controversial issues in contemporary bioethics (see Supplementary Table ??). We recruited a politically-balanced sample (N = `r length(unique(pretest$prolific))`) of U.K. residents via Prolific. Each participant was presented with all 30 statements in a randomized order and asked to report their agreement or disagreement using a slider. On the basis of this pretest, we selected pairs of normative/factual statements (one concerning the same issue), by applying three criteria: 

```{r}
tbl <- table1(pretest)

ft <- flextable::flextable(tbl)
ft <- merge_v(ft, j = ~ Issue + `Fact-Norm correlation`)
theme_vanilla(ft)
```
## Study 1a: Consequentialist Reflection
In this exploratory study, we examined the effect of reflection by asking study participants to reflect on a given issue and produce a short argumentative text. We employed a set of normative/factual statement pairs regarding 7 notable, possibly morally controversial issues. 
We examined the effect of reflection by asking participants to express (on a closed scale) their attitude towards each of the 7 normative statements twice. Before the second time, they were also asked to ponder a factual statement (matching one of the normative statements) and write a short description of arguments they would use to support their view on that factual statement. 

```{r}
study1 <- read.csv('data/01_long_data.csv')
```

### Methods

`r nrow(study1[!duplicated(study1$PROLIFIC_PID),])` U.K. nationals recruited on Prolific completed both parts of the study (Age = `r round(mean(study1[!duplicated(study1$PROLIFIC_PID),]$age, na.rm=T),0)` years; `r sum(study1[!duplicated(study1$PROLIFIC_PID),]$gender==2)` women; `r sum(study1[!duplicated(study1$PROLIFIC_PID),]$RL=='Right')` participants declared right-wing political views).
The study consisted of two sessions. In the first session, participants were asked to express their attitude towards each of the seven factual or normative (depending on condition) statements using a 100-point slider labeled at both extremes and both tertiles. For each statement,  participants were also asked to express their confidence in the provided answer. They also answered a few basic demographic questions. 
After a seven-day delay, participants returned to take part in the second wave of the study. Each participant was presented with one randomly selected factual (Reflection condition) statement, and were asked to reflect for at least 45 seconds on what they believe and the reasons for believing it. Participants recorded their attitude on a 100-point soldier. They were informed that they would be asked to explain their point of view to others. On the next screen, participants were asked to explain their point of view in a text box. After that, participants were presented with the same 7 statements as in the first session and asked to express their attitudes to each of them.

### Analysis plan
Assuming that we would observe some effects of reflection, we also wanted to check if those effects are symmetric or asymmetric. 

Hypothesis 1. *An asymmetric outcome, in contrast, would be observed if the cognitive effort invested in reflection and/or rationalization overall shifts the population judgment in one direction, making it, say, more progressive rather than conservative.*

We calculated the difference between post- and pre-treatment assessments of a given participant for a given issue. As the measures for all items were rescaled to correlate with conservative views, a negative difference would indicate a post-treatment left-ward shift, while a positive difference would be associated with a rightward shift.

Hypothesis 2. *A possible symmetric setting is one in which the effect moderates extreme attitudes and beliefs. And so, reflecting on a factual statement regarding a given issue, people might realize they are uncertain on whether a given activity is (or is not) harmful, thus they might moderate their moral assessment of this issue.*

If the effect of our manipulation were to be symmetric (increased or decreased polarization), then we would expect to observe a change in the deviations from the median. Hence, we calculated the absolute deviation of every observation from the (grand) median for a given item. We then calculated the difference between post- and pre-treatment absolute deviations for each participant/item. The resulting measure would take a positive value if the manipulation resulted in an increased polarization and a negative value in the case of a moderation. 
Finally, we aimed to check whether the effects of reflection/rationalization are general or domain-specific. In the adopted design, we measured the change in beliefs and attitudes regarding 7 issues, while asking participants to perform the rationalization/reflection task only on one of them. This allowed us to check if any observed effect took place only with regard to the target issue, or also for the other 6 filler issues.
Results

Correlation Table. Norm1, Fact1, Norm2, Fact 2. 
Stability instability? Test retest
 specific asymmetric effect, by fitting mixed-effects models (with the difference measure as the dependent measure) on the subset of reflected/rationalized statements, with the random effect of issue. The model for the reflection condition rendered a negative intercept (b0 = -4.40, t(145) = -2.40, p = .018),
